Improved state-space models for inference about spatial and temporal variation in abundance from count data

### 2013-03-26 Dail_Mad_ext.tex
Trim intro and remove GWWA

\bibliographystyle{ecology}

@@

\textbf{Key words}: abundance, Dail and Madsen model, density-dependence,
Gompertz-logistic model, immigration, open population point count models,
random observer effects, range, Ricker-logistic model, zero-inflated

@@

A more serious problem than the ones associated with the Gaussian
assumptions is that the parameters of the simple state-space models
such as Eq~\ref{eq:ss1} are not identifiable in many
circumstances \citep{polansky_etal:2009}. Recall that the
Eq.~\ref{eq:ss1} did not specify distributions for $N_1$; % or $X_1$;
however, $N_1$ is a random variable and hence there is uncertainty
that should be accounted for.  
%are needed to fully specify the model.
To be more specific, a fully-specified state-space model requires at least
three probability distributions, which we represent
using bracket notation (i.e., $[Y|\Omega]$ is interpreted
as the probability distribution of the random variable $Y$ given the
parameter $\Omega$). The three probability distributions required for a
state-space model are:
%%%% JEFF, does this extra sentence help? This notation comes from
%%%% several of the original state-space papers and is widespread in
%%%% books like Andy's and Bill's.
%\hl{[I'm not familiar with bracket notation,
% so these equations are confusing to me. How widely used is this
% notation in ecology? Will others likely understand this? Also, we might want
%to save space by trimming a bunch of this discussion back down.]}
\begin{subequations}
  \label{eq:ss2}
  \begin{align}
    [N_1&|\bm{\theta}] \label{eq:ss2a} \\ \label{eq:ss2b}
[N_t&|N_{t-1},\bm{\Theta}] \quad \text{for} \; t=2,\hdots,T \\
\label{eq:ss2c}
    [X_t&|N_t,\bm{p}]  \qquad \; \text{for} \; t=1,\hdots,T
  \end{align}
\end{subequations}
where $\bm{\theta}$ are the process variation parameters for the
initial state, $\bm{\Theta}$ are the process variation parameters
influencing how abundance changes over time, and $\bm{p}$ are the
observation error parameters. As mentioned previously, Eq.~\ref{eq:ss2b}
and Eq.~\ref{eq:ss2c} are assumed to be Gaussian in classical state-space
models, but what should be the distribution for Eq.~\ref{eq:ss2a}?
%The three lines of general model
%described by Eq.~\ref{eq:ss2} correspond to the three equations shown
%for the specific example in Eq~\ref{eq:ss1}. However, in this case,
%we allow $N_1$ to be a random variable. But what should its
%istribution, $[N_1|\bm{\theta}]$ be?
And how could $\bm{\theta}$ be estimated given that there is only a single observation
available? Often, these problems are simply assumed away. For
example, sometimes researchers assume that there is no process variation in
the first year, although this seems unjustified. Alternatively, the
population may be assumed to be at equilibrium such that the expected
value of $N_t$ is constant through time. Although this 
makes the parameters $\bm{\theta}$ identifiable, assuming equilibrium
defeats the objective of many studies of population dynamics, namely
determining why a population varies over time.

@@

As with traditional state-space models, the DM model includes the
three conditionally related processes shown in Eq.~\ref{eq:ss2},
which correspond to (1) initial abundance, i.e. the
abundance at site $i$ during the first primary period,
denoted (2) abundance at time $t$ (for $t>1$) which depends upon
abundance at $t-1$, and (3) the
detection process \citep{dail_madsen:2011}.
The first two processes describe the state process---the
variation in abundance in space and time. The third process %simply 
describes the the relationship between
abundance and the observed count data.

@@

%Eqs.~\ref{eq:N1} and \ref{eq:Nt} fully specify the state model,
%i.e. the model for abundance dynamics. This state
%model has allowed us to replace the initial abundance
%distribution and transition models of conventional state-space models, with
%alternatives that (1) allow for modeling spatial variation in
%abundance, (2) respect the discrete nature of the counts, and
%(3) have parameters that are estimable.
The one
remaining task is to replace the observation model
$[X_t|N_t, \bm{p}]$ (Eq.~\ref{eq:ss2c}) with one that can be
more easily interpreted. The observation
model assumes that individuals are missed due to imperfect
detection. The simplest model for imperfect detection is
\begin{equation}
  X_{it} \sim \mathrm{Bin}(N_{it}, p)
  \label{eq:p1}
\end{equation}
%or equivalently,
%\begin{equation}
%[X_{it} | N_{it}, \bm{p}] = \binom{N_{it}}{X_{it}}
%\bm{p}^{X_{it}}(1-\bm{p})^{N_{it}-X_{it}}.
%  \label{eq:1}
%\end{equation}
Under this model, $\bm{p}$ is interpreted as the detection
probability parameter.

@@

We applied these models to North American Breeding Bird Survey
(BBS) data from 1966-2010 for two species in the bordering US states
Maryland and Virginia. For our focal species, we selected
ovenbirds (\textit{Seiurus aurocapilla}), an abundant and widespread
forest breeding migrant with a stable or increasing trend in the region \citep{porneluzi_etal:2011}
and golden-winged warblers (\textit{Vermivora chrysoptera}), which
only breeds in the western parts of these states and has been declining in the
region \citep{confer_etal:2011}.  %MORE INFORMATION ABOUT THE SPECIES?

@@

%HOW WE OBTAINED AND MANIPULATED PRISM DATA
We ran a series of maximum-likelihood based models for each
species, and then a series of Bayesian models. We started by testing
three models of initial abundance (Poisson, negative binomial, and
zero-inflated Poisson) with exponential dynamics
(Eq.~\ref{eq:exp}) and no covariates.  We selected the minimum Akaike's Information
Criterion (AIC) model from that set to test three additional
models for p: wind, first, and wind + first. We selected the minimum
AIC model from that set to test eight additional models of dynamics:
constant-recruitment, geometric-recruitment, Ricker-logistic, Gompertz-logistic, geometric-recruitment +
immigration, exponential + immigration, Ricker-logistic +
immigration, and Gompertz-logistic + immigration. %We then tested the effect of average
%minimum temperature for June and July on all dynamics parameters from
%the minimum AIC dynamics model.
As above, these models require a a finite upper bound ($N_{max}$) to integrate
over when run in a maximum likelihood
framework; we used 600 for ovenbirds and 350 for golden-winged warblers.

We ran the top ranked models from the maximum likelihood
analyses in a Bayesian framework.  We added random observer 
effects and environmental stochasticity (modeled as lognormal
variation in regional population growth rate, Eq.~\ref{eq:nuRand}) and, where appropriate, zero-inflated dynamics.  We used
non-informative priors. We tested for lack of convergence using
5 Markov chains for each model \citep{gelman_rubin:1992}.
%(Gelman and Rubin 1992).
For each chain we sampled the MCMC for at least 40,000 iterations, after at
least 3,000 adaptation and tuning samples.

We also used these models to obtain route and 
year specific abundance estimates.  Because of the memory requirements involved,
we obtained these estimates from a single chain.  We present these estimates for ovenbirds only.
%Latent inclusion factors for temperature effects?

@@

Estimates for the top ranked model for ovenbird
%(NB[$\Lambda (.) \alpha (.)$]Ricker-logistic+Immigration[$r(.)K(.) \iota(.)$]p(wind+1st)) 
were similar when
run in the Bayesian framework, except that the estimate of $r$
more than halved.
% (from 0.026 $\pm$ 0.006 to 0.011 $\pm$ 0.005 [for
%Bayesian model estimates we present mean and SD]). 
When random observer effects
were added, estimates for $\Lambda$ and $K$ increased dramatically,
%(from 31.6 $\pm$ 4.9 to 42.5 $\pm$ 7.3 and from 56.2 $\pm$ 18.2 to 101.8 $\pm$ 21.1,
%respectively)
and the estimate for r was intermediate. The %(0.018 $\pm$ 0.006)
estimate of the intercept for $p$ (on the logit scale) dropped. % from -1.5 $\pm$
%0.1 to -2.0 $\pm$ 0.1, and the estimate of $\sigma_p$ was 0.35 $\pm$ 0.03.  
When environmental stochasticity was added, the estimate for $K$ increased. %(to 112 $\pm$ 58.8) and $r$ was
%again intermediate (0.020 $\pm$ 0.010).  Adding environmental stochasticity alone had little effect
%on estimate of $p$ or $\Lambda$. 
The highest estimates of $\Lambda$ and $K$ came 
from the model that included environmental stochasticity and random observer effects %(43.8 $\pm$
%7.9 and 141.5 $\pm$ 69.5, respectively).  
Gelman and Rubin diagnostics and visual examination of the chain trajectory and density plots
provided some evidence for a lack of convergence in estimates of the
intercept for $p$ for all Bayesian models ran.

@@

There was slightly more support for the ZIP distribution for initial
abundance of golden-winged warbler than for the negative binomial; both were strongly
supported over the Poisson (Table 3A). The best
supported model for $p$ was an effect of first run (Table 3B, model
B.1). The best supported dynamics model was exponential
(Table 3C, model C.1). %; the estimate of r from this model was -0.058
%$\pm$ 0.017 (SE). There was no support for immigration added to the
%exponential or geometric-recruitment models (Table 3C, models C.4 and
%C.6).
%However, the Gompertz-logistic + immigration and Ricker-logistic + immigration
%models were supported over the corresponding models without immigration
%(Table 3C, models C.4, C.5, C.7, and C.8), despite low estimates
%of $\iota$ from these models (0.003 $\pm$ 0.003 and 0.003 $\pm$
%0.002, respectively).  
%There was considerable support for an effect of
%average minimum June and July temperature on r in the population
%growth model (Table 1B, model B.17), with a negative slope.
Because of the closeness of rankings for the ZIP and negative binomial
distributions for golden-winged warbler, we also ran the subsequent models with the
negative binomial initial abundance; rankings of subsequent
models were similar. %, but with a $p$ covariate (first) the negative
%binomial distribution outranked the ZIP.

Estimates for the top ranked model for golden-winged warbler
%(ZIP[Λ(.)α(.)]Exponential[r(.)]p(1st))
were similar when run in the Bayesian framework. When random
observer effects were added, the estimate for $\Lambda$ increased dramatically,
and the estimate of the intercept for $p$ dropped.
%(from 22.4 $\pm$ 7.6 to 55.5 $\pm$ 30.9). The estimate of the intercept
%for $p$ (on the logit scale) dropped from -3.5 $\pm$ 0.2 to -5.6 $\pm$ 1.1, and the
%estimate of $\sigma_p$ was 1.7 $\pm$ 0.7. 
Adding environmental stochasticity also increased the estimate of $\Lambda$ 
%(to 42.9 $\pm$ 17.9) 
and the model with both random effects had the highest estimate of $\Lambda$.
% (74.6 $\pm$ 32.9). 
Adding zero-inflation to the dynamics (as well as
the initial abundance) has no effect for exponential models, because populations
that started at 0 would stay at 0 in any case. Therefore, we tested
zero-inflated dynamics on the Exponential + Immigration model. Adding zero-inflation to the dynamics
increased the estimate of $\iota$, but had only minor % from 0 to 0.091 $\pm$ 0.049
effects on the other parameters. Gelman and Rubin diagnostics and visual
examination of the chain trajectory and density plots provided evidence for a lack of convergence in
estimates of $\Lambda$ for all Bayesian models ran, and for most parameters when random observer
effects were included.

@@

For both species, constant-recruitment dynamics models estimated very high
survival probabilities ($\omega$ = 1 $\pm$ 1.3e-05 for ovenbird and
0.935 $\pm$ 0.011 for
golden-winged warbler). The geometric-recruitment and geometric-recruitment + immigration
models estimated very low survival probabilities for ovenbird ($\omega$ =
0.058 $\pm$ 0.046 and 0.026 $\pm$ 0.060, respectively), 
but not for golden-winged warbler ($\omega$ = 0.715 $\pm$ 0.288 for both models).

@@

Our work has highlighted the limitations of classical state-space
models as applied to ecological time series data, and we have
demonstrated how a class of open population $N$-mixture models
proposed by \citet{dail_madsen:2011} overcomes these
limitations. We extended their model in several important
ways to accomodate common features of ecological
datasets, including sparse counts, zero-inflation, demographic
stochasticity, and random variation in observation error.
%We made four important developments of the open population
%N-mixture model proposed by \citet{dail_madsen:2011} that make the
%model more applicable to long-term data such as is collected
%by the North American Breeding Bird Survey.
%First, we have
%reconciled the objectives of traditional state-space models
%with open population $N$-mixture models by illustrating how
%classical models of population dynamics can be embedded within
%the framework. Second, we have demonstrated methods of
%accounting for zero-inflation in the time-series. Third, we
%have illustrated how additional random effects such as
%observer-specific detection probabilities can be
%accommodated. Fourth, we presented a Bayesian analysis of the
%model, which, makes it possible to incorporate prior information when
%available, and in some cases, facilitates parameter estimation.
We also demonstrated how many of the objectives of conventional
state-space modeling can be accomplished within this expanded
framework. For example, one of the primary aims of our paper was to demonstrate how classical
population growth models can be embedded in the DM model.
Although we view this as an important connection to make between the two
different classes of state-space models, we believe that
%In some ways,
such population growth models are more phenomenological
than mechanistic, and this runs counter to the motivation for
the DM model. However, estimating
demographic parameters from count data is an ambitious goal, and
the required assumption will not be valid in many cases,
especially when immigration and emigration occur. This was supported by the
unrealistic estimates of vital rates reported here and by the
lack of support for the mechanistic models relative to the population
growth models.

One strategy for adhereing to the original formulation of the
model that included survival and recruitment parameters would be to
use either prior information or additional data. For example, in a
Bayesian analysis existing estimates of demographic parameters
could be used as informative priors, which might make it possible to
seperate the effects of movement versus births and deaths.
Alternatively, ecologists could combine count data with data
from, for example,
capture-recapture data. Models that are fitted to count data and
demographic data are often referred to as integrated population models
\citep[IPM;][]{besbeas_etal:2002,
buckland_etal:2004,schaub_etal:2007}
and we believe that a DM model could work well in this context.
In fact, the DM model could be viewed as a specific case of the
IPM model proposed by \citet{buckland_etal:2004}.
However, we caution that when count data are affected by
movement and vital rates, it would be unwise to ignore the movement
processes.

@@

Even though the DM models and our extentions are conceptually simple, even in
their most stripped down form they include many random effects (the $N$'s, which
incorporate demographic stochasticity), making it challenging to compute the likelihood 
or implement MCMC algorithms. Nonetheless, in some cases additional random effects
may be of interest.  We have shown a few approaches for incorporating some of these
(environmental stochasticity and random observer effects).  Other approaches to modeling
each of these may also be worth trying.  Incorporating these random effects can increase
the difficulty of achieving model convergence, but they are certainly important.
For example, the estimate of carrying capacity for the ovenbird increased by more than 150\%
when environmental stochasticity and random observer effects were included.

@@

We have presented a test case for these models using data from
the BBS for two species. We demonstrate that these models can be
used to model and estimate initial abundance, detection
probability, and density-independent and dependent functions for
population dynamics. We tested these models in both the
frequentist and Bayesian frameworks and found strengths and
weaknesses of each approach. The Bayesian analyses permit more
flexibility and allow the incorporation of random observer
effects and zero-inflated dynamics. Our results suggest ignoring
variation between observers can lead to underestimation of
initial abundance and equilibrium population sizes. On the other
hand, our Bayesian analyses did have convergence problems,
particularly for the species (golden-winged warbler) with sparser data. It's
likely that poor model fits are related to the low estimates of
detection probability, which in turn are probably caused by the
large 400 m radius used for BBS point counts. Again, we believe
that direct information on detectability (such as repeated
visits within a season) would greatly improve the fit of these
models in both frameworks.  We also were able to obtain year
and site-specific estimates of abundance (Fig.~\ref{fig:oven_N}),
although with generally lower precision than more global estimates.  These sort of estimates
may be more reliable when abundance and dynamics are modeled 
as functions of habitat covariates, which we did not do here.
\begin{comment}
[Alternatively, could put discussions of our results where
they fit in by subject area, rather than separately. For
instance, could discuss BBS population dynamics model selection
in population models paragraph, BBS Bayesian vs.
frequentist strengths and weakness in the Bayesian paragraph, 
and golden-winged warbler ZIP results in ZIP
paragraph.]
\end{comment}

>>>>

\bibliographystyle{ecology}
\hyphenpenalty=100000

@@

\textbf{Key words}: abundance, Dail and Madsen model, density-dependence,
detectability, Gompertz-logistic model, immigration, open population point 
count models, random observer effects, Ricker-logistic model, zero-inflated

@@

A more serious problem %than the ones associated with the Gaussian assumptions 
is that the parameters of the simple state-space models
such as Eq~\ref{eq:ss1} are not identifiable in many
circumstances \citep{polansky_etal:2009}. Recall that the
Eq.~\ref{eq:ss1} did not specify distributions for $N_1$; % or $X_1$;
however, $N_1$ is a random variable and hence there is uncertainty
that should be accounted for.  
%are needed to fully specify the model.
%To be more specific, a fully-specified state-space model requires at least
%three probability distributions, which we represent
%using bracket notation (i.e., $[Y|\Omega]$ is interpreted
%as the probability distribution of the random variable $Y$ given the
%parameter $\Omega$). The three probability distributions required for a
%state-space model are:
%%%% JEFF, does this extra sentence help? This notation comes from
%%%% several of the original state-space papers and is widespread in
%%%% books like Andy's and Bill's.
%\hl{[I'm not familiar with bracket notation,
% so these equations are confusing to me. How widely used is this
% notation in ecology? Will others likely understand this? Also, we might want
%to save space by trimming a bunch of this discussion back down.]}
%\begin{subequations}
%  \label{eq:ss2}
%  \begin{align}
%    [N_1&|\bm{\theta}] \label{eq:ss2a} \\ \label{eq:ss2b}
%[N_t&|N_{t-1},\bm{\Theta}] \quad \text{for} \; t=2,\hdots,T \\
%\label{eq:ss2c}
%    [X_t&|N_t,\bm{p}]  \qquad \; \text{for} \; t=1,\hdots,T
%  \end{align}
%\end{subequations}
%where $\bm{\theta}$ are the process variation parameters for the
%initial state, $\bm{\Theta}$ are the process variation parameters
%influencing how abundance changes over time, and $\bm{p}$ are the
%observation error parameters. As mentioned previously, Eq.~\ref{eq:ss2b}
%and Eq.~\ref{eq:ss2c} are assumed to be Gaussian in classical state-space
%models, but what should be the distribution for Eq.~\ref{eq:ss2a}?
%The three lines of general model
%described by Eq.~\ref{eq:ss2} correspond to the three equations shown
%for the specific example in Eq~\ref{eq:ss1}. However, in this case,
%we allow $N_1$ to be a random variable. But what should its
%istribution, $[N_1|\bm{\theta}]$ be?
%And how could $\bm{\theta}$ be estimated given that there is only a single observation
%available? Often, these problems are simply assumed away. For
%example, sometimes researchers assume that there is no process variation in
%the first year, although this seems unjustified. Alternatively, the
%population may be assumed to be at equilibrium such that the expected
%value of $N_t$ is constant through time. Although this 
%makes the parameters $\bm{\theta}$ identifiable, assuming equilibrium
%defeats the objective of many studies of population dynamics, namely
%determining why a population varies over time.

@@

As with traditional state-space models, the DM model includes the
three conditionally related processes shown in Eq.~\ref{eq:ss1},
which correspond to (1) initial abundance, i.e. the
abundance at site $i$ during the first primary period,
denoted (2) abundance at time $t$ (for $t>1$) which depends upon
abundance at $t-1$, and (3) the
detection process \citep{dail_madsen:2011}.
The first two processes describe the state process---the
variation in abundance in space and time. The third process %simply 
describes the the relationship between
abundance and the observed count data.

@@

%Eqs.~\ref{eq:N1} and \ref{eq:Nt} fully specify the state model,
%i.e. the model for abundance dynamics. This state
%model has allowed us to replace the initial abundance
%distribution and transition models of conventional state-space models, with
%alternatives that (1) allow for modeling spatial variation in
%abundance, (2) respect the discrete nature of the counts, and
%(3) have parameters that are estimable.
The one
remaining task is to replace the observation model
$X_t = N_t + \zeta_t$ (Eq.~\ref{eq:ss1c}) with one that can be
more easily interpreted. The observation
model assumes that individuals are missed due to imperfect
detection. The simplest model for imperfect detection is
\begin{equation}
  X_{it} \sim \mathrm{Bin}(N_{it}, p)
  \label{eq:p1}
\end{equation}
%or equivalently,
%\begin{equation}
%[X_{it} | N_{it}, \bm{p}] = \binom{N_{it}}{X_{it}}
%\bm{p}^{X_{it}}(1-\bm{p})^{N_{it}-X_{it}}.
%  \label{eq:1}
%\end{equation}
Under this model, $\bm{p}$ is interpreted as the detection
probability parameter.

@@

We applied these models to North American Breeding Bird Survey
(BBS) data from 1966-2010 in the bordering US states
Maryland and Virginia. For our focal species, we selected
ovenbirds (\textit{Seiurus aurocapilla}), an abundant and widespread
forest breeding migrant with a stable or increasing trend in the region \citep{porneluzi_etal:2011}.
%and golden-winged warblers (\textit{Vermivora chrysoptera}), which
%only breeds in the western parts of these states and has been declining in the
%region \citep{confer_etal:2011}.  %MORE INFORMATION ABOUT THE SPECIES?

@@

%HOW WE OBTAINED AND MANIPULATED PRISM DATA
We ran a series of maximum-likelihood based models, 
and then a series of Bayesian models. We started by testing
three models of initial abundance (Poisson, negative binomial, and
zero-inflated Poisson) with exponential dynamics
(Eq.~\ref{eq:exp}) and no covariates.  We selected the minimum Akaike's Information
Criterion (AIC) model from that set to test three additional
models for p: wind, first, and wind + first. We selected the minimum
AIC model from that set to test eight additional models of dynamics:
constant-recruitment, geometric-recruitment, Ricker-logistic, Gompertz-logistic, geometric-recruitment +
immigration, exponential + immigration, Ricker-logistic +
immigration, and Gompertz-logistic + immigration. %We then tested the effect of average
%minimum temperature for June and July on all dynamics parameters from
%the minimum AIC dynamics model.
As above, these models require a a finite upper bound ($N_{max}$) to integrate
over when run in a maximum likelihood
framework; we used 600. % for ovenbirds and 350 for golden-winged warblers.

We ran the top ranked models from the maximum likelihood
analyses in a Bayesian framework.  We added random observer 
effects and environmental stochasticity (modeled as lognormal
variation in regional population growth rate, Eq.~\ref{eq:nuRand}). % and, where appropriate, zero-inflated dynamics.  
We used
non-informative priors. We tested for lack of convergence using
5 Markov chains for each model \citep{gelman_rubin:1992}.
%(Gelman and Rubin 1992).
For each chain we sampled the MCMC for at least 40,000 iterations, after at
least 3,000 adaptation and tuning samples.

We also used these models to obtain route and 
year specific abundance estimates.  Because of the memory requirements involved,
we obtained these estimates from a single chain.  %We present these estimates for ovenbirds only.
%Latent inclusion factors for temperature effects?

@@

Estimates for the top ranked model for ovenbird
%(NB[$\Lambda (.) \alpha (.)$]Ricker-logistic+Immigration[$r(.)K(.) \iota(.)$]p(wind+1st)) 
were similar when
run in the Bayesian framework, except that the estimate of $r$
more than halved.
% (from 0.026 $\pm$ 0.006 to 0.011 $\pm$ 0.005 [for
%Bayesian model estimates we present mean and SD]). 
When random observer effects
were added, estimates for $\Lambda$ and $K$ increased dramatically,
%(from 31.6 $\pm$ 4.9 to 42.5 $\pm$ 7.3 and from 56.2 $\pm$ 18.2 to 101.8 $\pm$ 21.1,
%respectively)
and the estimate for r was intermediate. The %(0.018 $\pm$ 0.006)
estimate of the intercept for $p$ (on the logit scale) dropped. % from -1.5 $\pm$
%0.1 to -2.0 $\pm$ 0.1, and the estimate of $\sigma_p$ was 0.35 $\pm$ 0.03.  
When environmental stochasticity was added, the estimate for $K$ increased. %(to 112 $\pm$ 58.8) and $r$ was
%again intermediate (0.020 $\pm$ 0.010).  Adding environmental stochasticity alone had little effect
%on estimate of $p$ or $\Lambda$. 
The highest estimates of $\Lambda$ and $K$ came 
from the model that included environmental stochasticity and random observer effects. %(43.8 $\pm$
%7.9 and 141.5 $\pm$ 69.5, respectively).  
Gelman and Rubin diagnostics and visual examination of the chain trajectory and density plots
provided some evidence for a lack of convergence in estimates of the
intercept for $p$ for all Bayesian models ran.

@@

%There was slightly more support for the ZIP distribution for initial
%abundance of golden-winged warbler than for the negative binomial; both were strongly
%supported over the Poisson (Table 3A). The best
%supported model for $p$ was an effect of first run (Table 3B, model
%B.1). The best supported dynamics model was exponential
%(Table 3C, model C.1). %; the estimate of r from this model was -0.058
%$\pm$ 0.017 (SE). There was no support for immigration added to the
%exponential or geometric-recruitment models (Table 3C, models C.4 and
%C.6).
%However, the Gompertz-logistic + immigration and Ricker-logistic + immigration
%models were supported over the corresponding models without immigration
%(Table 3C, models C.4, C.5, C.7, and C.8), despite low estimates
%of $\iota$ from these models (0.003 $\pm$ 0.003 and 0.003 $\pm$
%0.002, respectively).  
%There was considerable support for an effect of
%average minimum June and July temperature on r in the population
%growth model (Table 1B, model B.17), with a negative slope.
%Because of the closeness of rankings for the ZIP and negative binomial
%distributions for golden-winged warbler, we also ran the subsequent models with the
%negative binomial initial abundance; rankings of subsequent
%models were similar. %, but with a $p$ covariate (first) the negative
%binomial distribution outranked the ZIP.

%Estimates for the top ranked model for golden-winged warbler
%%(ZIP[Λ(.)α(.)]Exponential[r(.)]p(1st))
%were similar when run in the Bayesian framework. When random
%observer effects were added, the estimate for $\Lambda$ increased dramatically,
%and the estimate of the intercept for $p$ dropped.
%%(from 22.4 $\pm$ 7.6 to 55.5 $\pm$ 30.9). The estimate of the intercept
%%for $p$ (on the logit scale) dropped from -3.5 $\pm$ 0.2 to -5.6 $\pm$ 1.1, and the
%%estimate of $\sigma_p$ was 1.7 $\pm$ 0.7. 
%Adding environmental stochasticity also increased the estimate of $\Lambda$ 
%%(to 42.9 $\pm$ 17.9) 
%and the model with both random effects had the highest estimate of $\Lambda$.
%% (74.6 $\pm$ 32.9). 
%Adding zero-inflation to the dynamics (as well as
%the initial abundance) has no effect for exponential models, because populations
%that started at 0 would stay at 0 in any case. Therefore, we tested
%zero-inflated dynamics on the Exponential + Immigration model. Adding zero-inflation to the dynamics
%increased the estimate of $\iota$, but had only minor % from 0 to 0.091 $\pm$ 0.049
%effects on the other parameters. Gelman and Rubin diagnostics and visual
%examination of the chain trajectory and density plots provided evidence for a lack of convergence in
%estimates of $\Lambda$ for all Bayesian models ran, and for most parameters when random observer
%effects were included.

@@

%For both species, 
Constant-recruitment dynamics models estimated very high
survival probabilities ($\omega$ = 1 $\pm$ 1.3e-05). % for ovenbird and
%0.935 $\pm$ 0.011 for
%golden-winged warbler). 
The geometric-recruitment and geometric-recruitment + immigration
models estimated very low survival probabilities %for ovenbird 
($\omega$ =
0.058 $\pm$ 0.046 and 0.026 $\pm$ 0.060, respectively). 
%but not for golden-winged warbler ($\omega$ = 0.715 $\pm$ 0.288 for both models).

@@

Our work has highlighted the limitations of classical state-space
models as applied to ecological time series data, and we have
demonstrated how a class of open population $N$-mixture models
proposed by \citet{dail_madsen:2011} overcomes these
limitations. We extended their model in several important
ways to accomodate common features of ecological
datasets, including sparse counts, zero-inflation, environmental
stochasticity, and random variation in observation error.
%We made four important developments of the open population
%N-mixture model proposed by \citet{dail_madsen:2011} that make the
%model more applicable to long-term data such as is collected
%by the North American Breeding Bird Survey.
%First, we have
%reconciled the objectives of traditional state-space models
%with open population $N$-mixture models by illustrating how
%classical models of population dynamics can be embedded within
%the framework. Second, we have demonstrated methods of
%accounting for zero-inflation in the time-series. Third, we
%have illustrated how additional random effects such as
%observer-specific detection probabilities can be
%accommodated. Fourth, we presented a Bayesian analysis of the
%model, which, makes it possible to incorporate prior information when
%available, and in some cases, facilitates parameter estimation.
We also demonstrated how many of the objectives of conventional
state-space modeling can be accomplished within this expanded
framework. For example, one of the primary aims of our paper was to demonstrate how classical
population growth models can be embedded in the DM model.
Although we view this as an important connection to make between the two
different classes of state-space models, we believe that
%In some ways,
such population growth models are more phenomenological
than mechanistic, and this runs counter to the motivation for
the DM model. However, estimating
demographic parameters from count data is an ambitious goal, and
the required assumption will not be valid in many cases,
especially when immigration and emigration occur. This was supported by the
unrealistic estimates of vital rates reported here and by the
lack of support for the mechanistic models relative to the population
growth models.

One strategy for adhereing to the original formulation of the
model that included survival and recruitment parameters would be to
use either prior information or additional data. For example, in a
Bayesian analysis existing estimates of demographic parameters
could be used as informative priors, which might make it possible to
seperate the effects of movement versus births and deaths.
Alternatively, ecologists could combine count data with data
from, for example,
capture-recapture data. Models that are fitted to count data and
demographic data are often referred to as integrated population models
\citep[IPM;][]{besbeas_etal:2002,
buckland_etal:2004,schaub_etal:2007}
and we believe that a DM model could work well in this context.
In fact, the DM model could be viewed as a specific case of the
IPM proposed by \citet{buckland_etal:2004}.
However, we caution that when count data are affected by
movement and vital rates, it would be unwise to ignore the movement
processes.

@@

Even though the DM models and our extentions are conceptually simple, even in
their most stripped down form they include many random effects (the $N$'s, which
incorporate demographic stochasticity), making it challenging to compute the likelihood 
or implement MCMC algorithms. Nonetheless, in some cases additional random effects
may be of interest.  We have shown a few approaches for incorporating some of these
(environmental stochasticity and random observer effects).  Other approaches to modeling
each of these may also be worth trying.  Incorporating these random effects can increase
the difficulty of achieving model convergence, but they are certainly important.
For example, the estimate of semi-equilibrium abundance for the ovenbird increased by more than 150\%
when environmental stochasticity and random observer effects were included.

@@

We have presented a test case for these models using data from
the BBS for ovenbirds. We demonstrate that these models can be
used to model and estimate initial abundance, detection
probability, and density-independent and dependent functions for
population dynamics. We tested these models in both the
frequentist and Bayesian frameworks and found strengths and
weaknesses of each approach. The Bayesian analyses permit more
flexibility and allow the incorporation of random observer
effects, environmental stochasticity, and zero-inflated dynamics. 
Our results suggest ignoring
variation between observers can lead to underestimation of
initial abundance and equilibrium population sizes. On the other
hand, our Bayesian analyses did have some convergence problems.
%particularly for the species (golden-winged warbler) with sparser data. 
It's likely that poor model fits are related to the low estimates of
detection probability, which in turn are probably caused by the
large 400 m radius used for BBS point counts. Again, we believe
that direct information on detectability (such as repeated
visits within a season) would greatly improve the fit of these
models in both frameworks.  We also were able to obtain year
and site-specific estimates of abundance (Fig.~\ref{fig:oven_N}),
although with generally lower precision than more global estimates.  These sort of estimates
may be more reliable when abundance and dynamics are modeled 
as functions of habitat covariates, which we did not do here.
\begin{comment}
[Alternatively, could put discussions of our results where
they fit in by subject area, rather than separately. For
instance, could discuss BBS population dynamics model selection
in population models paragraph, BBS Bayesian vs.
frequentist strengths and weakness in the Bayesian paragraph, 
and golden-winged warbler ZIP results in ZIP
paragraph.]
\end{comment}
